---
title: "Liquidity Risk"
subtitle: "FIN 331 Lecture 5"
author: 
  name: "Andrew Ainsworth"
  affiliation: "University of Wollongong"
format: 
  revealjs:
    theme: FIN331_css_L2.scss
    html-table-processing: none
    chalkboard: 
      buttons: false
      theme: whiteboard
      boardmarker-width: 2
      transition: 1
      #src: FIN331_chalkboard_L5.json
execute: 
  eval: true
  echo: false
editor: 
  markdown: 
    wrap: 72
---

## Lecture outline

- What is liquidity risk?
  - Liquidity risk can arise on both the asset and liability side of the balance sheet
  - Banks need to access funds to plug liquidity shortfall
- How can banks measure liquidity risk
  - Finance gap
  - Balance sheet ratios
  - Liquidity index
  - Liquidity coverage ratio
  - Net stable funding ratio
- Bank runs and solvency problems

\

- Reading
    - Ch 12


## What is liquidity risk?

- Liquidity reflects the ability to convert a security into cash in a short amount of time without having an adverse impact on the price of the security 

:::{.fragment}
- Banks face liquidity risk from both the asset and liability sides of the balance sheet
  - Liability-side liquidity risk: A significant unexpected increase in depositors demanding immediate access to cash can cause banks to have to liquidate assets to fund these withdrawals 
  - Asset-side liquidity risk: If borrowers with access to credit lines and loan commitments unexpectedly draw down on these funds
:::

:::{.fragment}
- It is normal for banks to face liquidity risk
- Banks are able to predict withdrawals based on past behaviour and seasonal factors 
- In extreme cases liquidity risk issues can lead to bank failure
:::


## What is liquidity risk? 

- Cash is the most liquid asset but it does not earn any interest 
  - In general, there is an inverse relationship between the liquidity of an asset and its return
- Banks try to minimize their cash holdings because they do not earn interest
- Banks invest in longer maturity, less liquid assets to earn interest revenue
- Liquid assets can be sold and converted into cash with relative ease and speed
- Less liquid assets can be difficult to sell and if the seller demands immediacy when selling, they will often have to accept a much lower price for the asset (fire-sale price)


## Liability-side liquidity risk

- A large proportion of the liabilities on a bank's balance sheet are deposits
- A substantial amount of these deposits are at-call meaning that they can be immediately withdrawn in cash
- However, only a small proportion of these deposits will be withdrawn on any given day
  - In this situation, deposits are a stable source of funds
- Any withdrawals can also be offset with new inflows of deposits
  - $\text{Net deposit drain = Deposit outflows - deposit inflows}$
- The inflow and outflow of deposits for a given bank is predictable on a "normal" day
- The bank has two source of funds if there is an unexpected increase in outflows of deposits (positive net deposit drain)
  - Borrow additional funds (purchased liquidity management)
  - Sell assets (stored liquidity management)


## Purchased liquidity management

- Banks can purchase liquidity in short-term money markets  
  - The [interbank overnight cash market](https://www.rba.gov.au/mkt-operations/domestic-market-ops-and-standing-facilities.html)
  - The [repurchase agreement market](https://www.rba.gov.au/publications/bulletin/2024/jul/the-australian-repo-market-a-short-history-and-recent-evolution.html) or "repo" market
- The bank could issue fixed-maturity wholesale certificates of deposit
- The bank is paying market rates for these funds, which are higher than the rate charged on deposits 
  - This can be a deterrent to purchased liquidity management
- A benefit of purchased liquidity management is that the asset side of the balance sheet is not impacted 
  - The adjustment to the liquidity event occurs on the liability side of the balance sheet


## Stored liquidity management

- Banks can sell some of the assets on its balance sheet to fund the net deposit drain 
  - Utilizing stored liquidity
- Banks can use cash reserves to meet the net deposit drain
- Both sides of the balance sheet contract when using stored liquidity management
- Stored liquidity management is also costly as banks need to hold cash, which earns a low, or zero, interest rate
- This creates an opportunity cost as the funds could have been invested in higher yielding assets than cash or lent out to borrowers
- Stored liquidity management and purchased liquidity management are not mutually exclusive and banks can use both approaches simultaneously


## Asset-side liquidity risk 

- Banks can provide a credit line or loan commitment to a borrower that allows them to borrow funds over a commitment period at a time of their choosing 
  - The borrower chooses the amount and timing that the funds are borrowed
- This can lead to unexpected liquidity demands from borrowers and give rise to liquidity risk

:::{.fragment}
- Banks face asset-side liquidity risk from their investment portfolio 
- If liquidity dries up in the markets where their investment assets trade then the value of assets can decline if banks need to sell assets quickly (fire-sale prices)
- This liquidity risk can be managed with either purchased liquidity management or stored liquidity management 
:::


## The financing gap

- The financing gap measures the difference between the value of loans and deposits

$$\text{Financing gap = Average loans - Average deposits}$$ 

:::{.fragment}
- Banks must use stored liquidity management or purchased liquidity management if the financing gap is positive

$$\text{Financing gap} = \text{Borrowed funds - Liquid assets}$$ 

:::

:::{.fragment}
- Banks will have higher borrowing if they have a larger financing gap and have fewer liquid assets to utilize
  - Higher borrowing can expose the bank to liquidity risk as it is relying on market interest rates and the supply of capital from lenders in these markets
  - Lenders in the money market could become concerned about the bank's creditworthiness and charge a credit risk premium or undertake credit rationing
  - This can lead to insolvency if the financing requirement cannot be met with liquid assets or borrowing
- An increasing financing gap can indicate future liquidity problems 
:::


## Peer goup ratio comparisons

- Liquidity risk can be identified through ratios 
- Loans to deposits ratio
  - This can indicate the reliance of the bank on purchased funds from the money market to fund loans
  - A high value could indicate that liquidity problems may arise if the bank is at or near its borrowing limit
- Loans to assets ratio 
  - This highlights a need for liquidity if the ratio is high as loans are longer-term and illiquid
- Unused commitments (or lines of credit) to assets ratio
  - This ratio can pick up asset-side liquidity risk if borrowers draw down on these commitments 


## Net loans to deposits

```{r}
library(tidyverse)
library(ggplot2)
library(scales)

gtable <- read_csv("gtable.csv")

gtable %>%
  filter(var == "Net loans to deposits", !is.na(value)) %>%
  filter(type == "All ADIs") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(linewidth = 0.7, colour="firebrick3") +
  labs(x = "Date", y = "Percent (%)") +
  theme(axis.title.x = element_text(size=12),
        axis.title.y = element_text(size=12),
        axis.text = element_text(size=12),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
        panel.grid.major.y = element_line(colour = "grey", linewidth = 0.5, linetype = "dotted"),
        text = element_text(family ="sans"),
        legend.title = element_blank(),
        legend.text = element_text(size=11)) +
  scale_y_continuous(expand = expansion(mult = 0), 
                     labels = scales::label_percent(),
                     limits = c(1, 1.2),
                     breaks = seq(1, 1.2, 0.02))
```

## Liquidity index

- This index measures potential losses a bank could face if they were forced to sell assets at fire-sale prices rather than fair value prices

$$I = \sum_{i=1}^N \left[ (w_i) \left( \frac{P_i}{P_i^*} \right) \right]$$

- $w_i$ is the percent allocation to each asset 
- $\sum_{i=1}^N w_i = 1$
- $0 \le I \le 1$
- $P_i$ are the fire-sale asset prices 
- $P_i^*$ are the fair market prices
- A larger difference between immediate fire-sale asset prices $(P_i)$ and fair market prices $(P_i^*)$ indicates that their portfolio is less liquid
- The liquidity index can be compared to peers

## Liquidity coverage ratio

- The [liquidity coverage ratio](https://www.bis.org/basel_framework/standard/LCR.htm) (LCR) measures the banks level of high-quality liquid assets (HQLA) that can be converted into cash to meet liquidity needs over a 30-day horizon
  - The measure incorporates a scenario of both institution-specific and systemic shocks
  - Does the bank have enough liquid assets to survive a 30-day liquidity crisis?
  - The LCR must remain above 100%

$$LCR = \frac{\text{Stock of HQLA}}{\text{Total net cash outflows over the next 30 days}}$$

:::{.fragment}
- HQLA are defined as follows:
  - Liquid assets must remain liquid in times of stress and be unencumbered 
  - Liquid assets are divided into level 1 and level 2
  - Traded in large, deep and active repo or cash markets characterized by a low level of concentration
  - Have a proven record as a reliable source of liquidity even during stressed market conditions
:::


## Liquidity coverage ratio

- Level 1 HQLA
  - No cap on amount 
  - Includes, for example, cash, central bank reserves, certain sovereign bonds and claims on other supranational agencies (e.g. International Monetary Fund)

:::{.fragment}
- Level 2 HQLA
  - Level 2 amount is capped at 40%
  - Level 2A assets
    - Includes certain claims on central banks, sovereigns, corporate debt issued by non-financial issuers rated above AA-
    - A minimum 15% haircut applied to the value of level 2A assets
  - Level 2B assets
    - Includes residential mortgage backed securities rated above AA, corporate debt rated above BBB-, equity shares from non-financial companies
    - Level 2B assets limited to 15% of HQLA
    - A minimum 25% haircut applied to the value of residential mortgage-backed securities
    - A minimum 50% haircut applied to the value of corporate bonds and equities
:::


## Liquidity coverage ratio

- Total net cash outflows is the denominator of the LCR
- Total net cash outflows take into account assumptions on the magnitude of the draw-down
  - For example, 5% of stable deposits are assumed to be withdrawn

$$\text{Total net cash outflows over the next 30 days} =\\ 
\text{Total expected cash outflows} -\\
\text{Min(Total expected cash inflows, 75% of total expected cash outflows)}$$

\

:::{.fragment}
- Not all ADIs in Australia are required to report the LCR
- APRA decides which banks need to meet the LCR requirement 
  - There are currently 12 banks that need to meet the LCR
:::


## Net stable funding ratio

- The net stable funding ratio (NSFR) examines liquidity over the entire balance sheet
- The NSFR encourages banks to to use longer-term stable sources of funding

$$ NSFR = \frac{\text{Available amount of stable funding}}{\text{Required amount of stable funding}}$$

:::{.fragment}
- Available stable funding (ASF) includes
  - Bank capital
  - Preferred stock with a maturity > 1 year
  - Liabilities with maturities > 1 year
  - The portion of deposits expected to remain during a period of idiosyncratic stress
- These sources of funds are given a haircut based on pre-determined values
:::

## Net stable funding ratio

- The required stable funding (RSF) amount is based on assumptions regarding the liquidity risk characteristics of assets, off-balance-sheet exposures, and other selected activities 
- The RSF takes account of assets and the amount of stable funding needed for that asset based on a predetermined percentage (RSF factor)
  - The asset definitions from the LCR are used
  - For example, corporate debt securities with a credit rating of between A+ and BBB- require a 50% stable funding rate
- The RSF factors are applied to each asset and represent the regulator's view on how much of the asset should be supported with stable funding
- The RSF factors are intended to approximate the amount of a particular asset that could not be sold or used as collateral in a secured borrowing during a severe liquidity event lasting one year



## Minimum liquidity holding

- APRA uses a minimum liquidity holdings (MLH) measure to monitor liquidity for smaller ADIs
- The MLH is defined in [Prudential Standard APS 210 Liquidity](https://www.legislation.gov.au/F2025L00653/latest/text)
- An MLH ADI must maintain a minimum holding of 9% of its liabilities in specified liquid asset
- Liquid assets include
  - Notes, coins and settlement funds
  - Australian government securities
  - Debt securities issued by supranationals and foreign governments
  - Bank bills, certificates of deposits (CDs) and debt securities issued by ADIs
  - Deposits that are at call or can be converted into cash within two business days


## Average LCR - Selected ADIs

```{r}
liqtable <- read_csv("liqtable.csv")

liqtable %>%
  filter(var == "Liquidity coverage ratio (LCR)", !is.na(value)) %>%
  filter(type == "LCR ADIs") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(linewidth = 0.7, colour="gold2") +
  labs(x = "Date", y = "Percent (%)") +
  theme(axis.title.x = element_text(size=12),
        axis.title.y = element_text(size=12),
        axis.text = element_text(size=12),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
        panel.grid.major.y = element_line(colour = "grey", linewidth = 0.5, linetype = "dotted"),
        text = element_text(family ="sans"),
        legend.title = element_blank(),
        legend.text = element_text(size=11)) +
  scale_y_continuous(expand = expansion(mult = 0), 
                     labels = scales::label_percent(),
                     limits = c(1.1, 1.5),
                     breaks = seq(1.1, 1.5, 0.05)) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y")
```

## Average NSFR - Selected ADIs

```{r}
liqtable <- read_csv("liqtable.csv")

liqtable %>%
  filter(var == "NSFR ratio", !is.na(value)) %>%
  filter(type == "LCR ADIs") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(linewidth = 0.7, colour="purple4") +
  labs(x = "Date", y = "Percent (%)") +
  theme(axis.title.x = element_text(size=12),
        axis.title.y = element_text(size=12),
        axis.text = element_text(size=12),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
        panel.grid.major.y = element_line(colour = "grey", linewidth = 0.5, linetype = "dotted"),
        text = element_text(family ="sans"),
        legend.title = element_blank(),
        legend.text = element_text(size=11)) +
  scale_y_continuous(expand = expansion(mult = 0), 
                     labels = scales::label_percent(),
                     limits = c(1.1, 1.3),
                     breaks = seq(1.1, 1.3, 0.02)) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y")
```


## Average MLH - Selected ADIs

```{r}
liqtable <- read_csv("liqtable.csv")

liqtable %>%
  filter(var == "MLH ratio", !is.na(value)) %>%
  filter(type == "MLH ADIs") %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(linewidth = 0.7, colour="sienna1") +
  labs(x = "Date", y = "Percent (%)") +
  theme(axis.title.x = element_text(size=12),
        axis.title.y = element_text(size=12),
        axis.text = element_text(size=12),
        panel.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
        panel.grid.major.y = element_line(colour = "grey", linewidth = 0.5, linetype = "dotted"),
        text = element_text(family ="sans"),
        legend.title = element_blank(),
        legend.text = element_text(size=11)) +
  scale_y_continuous(expand = expansion(mult = 0), 
                     labels = scales::label_percent(1),
                     limits = c(0.14, 0.21),
                     breaks = seq(0.14, 0.21, 0.01)) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y")
```

## Unexpected deposit drains and bank runs

- Banks can experience significant liquidity problems if there are large, unexpected deposit drains
  - Depositors might be concerned about the bank's solvency
  - The failure of a related bank can lead to contagion as depositors become concerned about the solvency of other banks

:::{.fragment}
- Demand deposits are paid on a first-come, first-served basis
  - A depositor will either get paid in full or receive nothing
  - It is rational for depositors to join the queue and withdraw funds
  - As a bank run increases in intensity, more depositors join the withdrawal queue and a liquidity crisis develops
:::

:::{.fragment}
- A bank can initially use stored liquidity or purchased liquidity but these sources will eventually be exhausted 
  - The cost of purchased funds will increase and counterparties will stop lending to a bank that is experiencing a run
- The bank will need to sell less liquid assets at fire-sale prices
:::

:::{.fragment}
- Ultimately, the liquidity issue becomes a solvency issue
:::


## Managing liquidity risk

- Banks can reduce liquidity risk by holding large amounts of cash and liquid assets (e.g. government bonds)
  - But they will receive lower yields on these assets 
- Less liquid assets provide a higher yield that reflects a higher credit risk premium and a lack of liquidity

:::{.fragment}
- Banks must determine the appropriate mix of lower-yielding, liquid assets versus higher-yielding, less liquid assets
- There is a tradeoff between risk and return for the bank
  - Highly liquid assets $\rightarrow$ low bank profitability but low liquidity risk
  - Less liquid assets $\rightarrow$ high bank profitability but high liquidity risk
- Remember that regulators are monitoring liquidity of assets and liabilities 
:::


## Managing liquidity risk

- Banks should also manage their liability structure to reduce the need for large amounts of liquid assets to meet liability withdrawals
- Excessive use of purchased funds in the liability structure can result in a liquidity crisis if investors lose confidence in the bank
  - Interest charged to the bank on money market borrowing will increase 
  - The market might refuse to roll over such funds

:::{.fragment}
- There is a trade-off between funding costs and liquidity
- The cheapest liabilities are those with the most liquidity risk
  - Demand deposits can be withdrawn immediately but pay a low interest rate
:::


## Conclusion

- Liquidity risk can arise from liabilities with depositors withdrawing funds
- Liquidity risk can arise from assets as borrowers draw down on loan commitments or banks investment portfolio is impacted by market movements
- Banks can use stored liquidity or purchased liquidity to manage liquidity shortfalls
- There are a number of liquidity risk measures that banks and regulators use
- Banks need to balance the liquidity of their assets with the liquidity risk of their liabilities
- Bank runs begin as a liquidity problem but can become a solvency problem
  - It is rational for depositors to withdraw funds in a bank run

\

- Next week: Sovereign Risk and Off-Balance Sheet Risk